### 2021-03-23 written/eLifeRevision/scaling_RNN_inference.tex
rnn3 figs and methods text

\thispagestyle{plain}
{\Large Interrogating theoretical models of neural computation with deep inference} \\
Sean R. Bittner$^{1}$, Agostina Palmigiano$^{1}$, Alex T. Piet$^{2,3,4}$, Chunyu A. Duan$^{5}$, Carlos D. Brody$^{2,3,6}$, \\
Kenneth D. Miller$^{1}$, and John P. Cunningham$^{7}$.

{\small
$^{1}$Department of Neuroscience, Columbia University, \\
$^{2}$Princeton Neuroscience Institute, \\
$^{3}$Princeton University, \\
$^{4}$Allen Institute for Brain Science, \\
$^{5}$Institute of Neuroscience, Chinese Academy of Sciences, \\
$^{6}$Howard Hughes Medical Institute, \\
$^{7}$Department of Statistics, Columbia University
}

\linenumbers
\section{Abstract}
%When analytic derivation of the relationship between model parameters and computational properties is intractable, approximate inference and simulation-based techniques are relied upon for scientific insight.
%We bring the use of deep generative models for probabilistic inference to bear on this problem, learning complex distributions of parameters that produce the specified properties of computation.
A cornerstone of theoretical neuroscience is the circuit model: a system of equations that captures a hypothesized neural mechanism.  
Such models are valuable when they give rise to an experimentally observed phenomenon -- whether behavioral or a pattern of neural activity -- and thus can offer insights into neural computation.
The operation of these mechanistic circuits, like all models, critically depends on the choices of model parameters.
A key process in neuroscientific modeling is then to identify the model parameters consistent with observed phenomena: to solve the inverse problem.
While statistical inference has proven effective on a broad variety of neuroscientific datasets, we clarify an important incongruity between theoretical approaches to neuroscience and this probabilistic methodology.
Theoretical neuroscience is focused on computational properties and how they emerge from biological mechanisms, rather than noisy experimental datasets and their quantified structure.
In this work, we present a novel technique to directly infer circuit model parameters producing these computational properties.
This method tailors deep inference, the use of deep neural networks for statistical inference, to the nature of theoretical inverse problems, enabling scaling to high dimensions and considerable technical simplifications.
With this method, we bring deep inference to bear on important questions in theoretical neuroscience, and demonstrate the broad range of insightful analyses this approach allows.
First, we emphasize the general applicability of this approach by inferring channel conductance parameters that produce a distinctive spiking frequency in a biophysical model of the stomatogastric ganglion.
Then, in a model of primary visual cortex with multiple neuron-types, where  analysis becomes untenable as more neuron-types are included, we use this technique to discover how noise properties govern excitatory population variability.
Next, in a model of superior colliculus, we identify and characterize two distinct regimes of connectivity that facilitate switching between opposite tasks throughout interleaved trials. 
Finally, we scale inference to 1,000-dimensional parameter spaces of RNN connectivity that exhibit stable, yet amplified responses.
These analyses illustrate how we can further leverage the power of deep learning towards solving inverse problems in theoretical neuroscience.

%-------- old sentences -----------

%Historically, the gold standard has been to analytically derive the relationship between model parameters and computational properties.  
%However, this enterprise quickly becomes infeasible as biologically realistic constraints are included into the model increasing its complexity, often resulting in \emph{ad hoc} approaches to understanding the relationship between model and computation.
%Importantly, the techniques we introduce offer a principled means to understand the implications of model parameter choices on computational properties of interest.  
%We then use it to go beyond linear theory of neuron-type input-responsivity in a model of primary visual cortex, gain a mechanistic understanding of rapid task switching in superior colliculus models, and attribute error to connectivity properties in recurrent neural networks solving a simple mathematical task.
%More generally, this work suggests a departure from realism vs tractability considerations, towards the use of modern machine learning for sophisticated interrogation of biologically relevant models.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}
The fundamental practice of theoretical neuroscience is to use a mathematical model to understand neural computation, whether that computation enables perception, action, or some intermediate processing.  
A neural circuit is systematized with a set of equations -- the model -- and these equations are motivated by biophysics, neurophysiology, and other conceptual considerations \cite{kopell1988coupled,  marder1998biophysics, abbott2008theoretical, wang2010neurophysiological}.
The function of this system is governed by the choice of model \emph{parameters}, which when configured in a particular way, give rise to a measurable signature of a computation.   
The work of analyzing a model then requires solving the inverse problem: given a computation of interest, how can we reason about particular parameter configurations?  
The inverse problem is crucial for reasoning about likely parameter values, uniquenesses and degeneracies, and predictions made by the model \cite{gutenkunst2007universally, o2014cell}.

Consider the idealized practice: one carefully designs a model and analytically derives how computational properties determine model parameters.
Seminal examples of this gold standard (which often adopt approaches from statistical physics) include our field's understanding of memory capacity in associative neural networks \cite{hopfield1982neural}, chaos and autocorrelation timescales in random neural networks \cite{sompolinsky1988chaos}, the paradoxical effect \cite{tsodyks1997paradoxical}, and decision making \cite{wong2006recurrent}.
Unfortunately, as circuit models include more biological realism, theory via analytical derivation becomes intractable.
Alternatively, we can gain insight into these complex models by identifying the distribution of parameters that produce a specific computational property.
By solving the inverse problem in this way, scientific analysis of complex biologically realistic models is made possible \cite{foster1993significance, prinz2004similar, achard2006complex, o2014cell, alonso2019visualization}.
%By solving the inverse problem in this way, scientists can reason about the sensitivity and robustness of the model with respect to different parameter combinations \cite{foster1993significance, prinz2004similar, achard2006complex, o2014cell, alonso2019visualization}.

%The preferred formalism for parameter identification in science is statistical inference, which has been used to great success in neuroscience through the stipulation of statistical generative models \cite{kass2001spike, brown1998statistical, paninski2004maximum, truccolo2005point, schneidman2006weak, druckmann2007novel, turner2007maximum, byron2009gaussian, macke2011empirical, park2011bayesian, granot2013stimulus, latimer2015single, lakshminarasimhan2018dynamic, duncker2019learning, ladenbauer2019inferring} (see review, \cite{paninski2018neural}). 
%However, most neural circuit models in theoretical neuroscience stipulate a noisy system of differential equations that can only be sampled or realized through forward simulation; they lack the explicit likelihood central to the probabilistic modeling toolkit.
%Therefore, the most popular approaches to the inverse problem have been likelihood-free methods such as approximate Bayesian computation (ABC) \cite{sisson2007sequential, liepe2014framework}, in which reasonable parameters are obtained via simulation and rejection.
% 
%Of course, the challenge of doing inference in complex models has arisen in many scientific fields.
%In response, the machine learning community has made remarkable progress in recent years, via the use of deep neural networks as powerful inference engines: a flexible function family that can map observations back to probability distributions quantifying the likely parameter configurations. 
%One celebrated example of this approach from machine learning, of which we draw key inspiration for this work, is the variational autoencoder (VAE) \cite{kingma2013auto, rezende2014stochastic}, which uses a deep neural network to induce an (approximate) posterior distribution on hidden variables in a latent variable model, given data. 
%Indeed, these tools have been used to great success in neuroscience as well, in particular for interrogating hidden states in models of both cortical population activity \cite{gao2016linear, zhao2017recursive, barello2018sparse, pandarinath2018inferring} and animal behavior \cite{wiltschko2015mapping, johnson2016composing, batty2019behavenet}. 
%These works have used deep neural networks to expand the domain of neural data sets amenable to statistical modeling \cite{paninski2018neural}.

One preferred formalism for parameter identification is statistical inference, which has been used to great success in neuroscience through the stipulation of statistical generative models \cite{kass2001spike, brown1998statistical, paninski2004maximum, truccolo2005point, schneidman2006weak, druckmann2007novel, turner2007maximum, byron2009gaussian, macke2011empirical, park2011bayesian, granot2013stimulus, latimer2015single, lakshminarasimhan2018dynamic, duncker2019learning, ladenbauer2019inferring} (see review, \cite{paninski2018neural}). 
Recent work has used variational autoencoders (VAEs) \cite{kingma2013auto, rezende2014stochastic} to interrogate hidden states in models of both cortical population activity \cite{gao2016linear, zhao2017recursive, barello2018sparse, pandarinath2018inferring} and animal behavior \cite{wiltschko2015mapping, johnson2016composing, batty2019behavenet}, thus expanding the domain of neural data sets amenable to statistical modeling.
However, most neural circuit models in theoretical neuroscience are noisy systems of differential equations that can only be sampled or realized through forward simulation; they lack the explicit likelihood necessary for statistical inference.
Therefore, the most popular approaches to theoretical inverse problems have been likelihood-free inference (LFI) methods \cite{sisson2007sequential, liepe2014framework}, in which reasonable parameters are obtained via simulation and rejection.
A flourishing new class of techniques \cite{gonccalves2019training, papamakarios2019sequential, hermans2020likelihood} use deep learning to improve upon traditional LFI approaches.
However, as we detail, all of these approaches require good datasets for the scientific question at hand.

%
%Existing approaches to the inverse problem in theoretical neuroscience fall short in three key ways. 
%First, theoretical models of neural computation aim to reflect a complex biological reality, and as a result, such models lack tractable likelihoods.
%Without an efficient calculation of the probability of model properties given model parameters, neuroscientists resort to approximate Bayesian computation \cite{beaumont2002approximate, marjoram2003markov, sisson2007sequential}, which requires a rejection heuristic, scales poorly, and only produces sets of accepted parameters lacking probabilities.
%Second, there is an undesirable trade-off between the flexibility and sampling speed of approximated posterior distributions.
%Sampling-based inference approaches (e.g. ABC and Markov chain Monte Carlo (MCMC) \cite{hastings1970monte, metropolis1953equation}) confer flexible approximations, yet scale poorly in number of parameters.
%While variational inference (VI) \cite{saul1998mean} often results in fast posterior sampling, existing practice relies heavily on simplified classes of distributions \cite{rezende2015variational}.  
%Third, such parameter inference methods are designed to operate on experimentally collected data-sets.
%Ultimately, the objects of interest in theoretical neuroscience are phenomena or features of the model rather than singular data-sets.

%To address these three challenges, we developed an inference methodology -- `emergent property inference' -- which learns a distribution over parameter configurations in a theoretical model.  
%This distribution has two critical properties: \emph{(i)} it is chosen such that draws from the distribution (parameter configurations) correspond to systems of equations that give rise to a specified emergent property (a set of constraints); and \emph{(ii)} it is chosen to have maximum entropy given those constraints, such that we identify all likely parameters and can use the distribution to reason about parametric sensitivity and degeneracies \cite{transtrum2015perspective}.  
%First, we use stochastic gradient techniques in the spirit of likelihood-free variational inference \cite{tran2017hierarchical} to enable inference in likelihood-free models of neural computation.
%Second, we stipulate a bijective deep neural network that induces a flexible family of probability distributions over model parameterizations with a probability density we can calculate \cite{rezende2015variational, dinh2017density, papamakarios2017masked}, which confers fast sampling and sensitivity measurements.
%Third, we quantify the notion of emergent properties as a set of moment constraints on datasets generated by the model.  
%Thus, an emergent property is not a single data realization, but a phenomenon or a feature of the model.
%Conditioning on an emergent property requires a variant of deep probabilistic inference methods, which we have previously introduced \cite{loaiza2017maximum}.
%Taken together, emergent property inference (EPI) provides a methodology for inferring parameter configurations consistent with a particular emergent phenomena in theoretical models.
%We use a classic example of parametric degeneracy in a biological system, the stomatogastric ganglion \cite{goldman2001global}, to motivate and clarify the technical details of EPI.

This work seeks to clarify an important incongruity between theoretical approaches to neuroscience and existing statistical inference methodology.
In theoretical neuroscience, we are concerned with the computational properties -- the \emph{emergent phenomena} -- of our models \cite{hopfield1982neural, sompolinsky1988chaos, tsodyks1997paradoxical, wong2006recurrent}, not noisy observed datasets \cite{paninski2018neural}.
To use the aforementioned inference paradigm, scientists must shoehorn such mathematical criteria into an artificial dataset compatible with existing statistical approaches.
Theorists are therefore barred from using the probabilistic modeling toolkit for science, unless they reformulate their inverse problem to fit an evidence accumulation framework.

These challenges motivate the development of a novel inference framework called emergent property inference (EPI).
As an adaption of variational inference \cite{saul1998mean}, EPI infers parameter distributions that produce an emergent property: not a singular dataset, but a collection of datasets  exhibiting some mathematical criteria.
EPI fixates on the inverse problem by constraining the predictions of the inferred parameter distribution to produce the emergent property exactly.
Conditioning on an emergent property requires a variant of deep probabilistic inference methods, which we have previously introduced \cite{loaiza2017maximum}.
By using deep probability distributions EPI flexibly captures parametric structure in mechanistic models.
This technique is designed to identify the full parameter space producing an emergent property and facilitate the seemless structural analysis of the inferred parameter distribution.
Thus, EPI provides neuroscientists with an efficient, versatile probabilistic modeling toolkit designed explicitly for theoretical inverse problems.

Equipped with this method, we bring deep inference to bear on theoretical neuroscience to an unprecedented extent.
Throughout this work, we showcase the capabilities of EPI on four neural circuit models across ranges of biological realism, neural system function, and network scale.
First, we show EPI's ability to capture subtle, nonlinear parametric structure in a stomatogastric ganglion subcircuit model \cite{gutierrez2013multiple}.
In a model of primary visual cortex \cite{palmigiano2020structure}, we show how to gain insight from multiple inferred distributions.
Next, we used EPI to idenitfy and structurally characterize multiple parametric regimes of superior colliculus activity in a model of task switching \cite{duan2018collicular}.
Finally, we emphasize the superior scalability of EPI compared to other LFI techniques by inferring high-dimensional distributions of RNN connectivities that exhibit amplified, yet stable responses -- a hallmark of cortical sensory systems \cite{more, bondanelli2020coding}.

Most importantly of all in this work, we present novel theories of neural computation borne from EPI analysis.
We identified an unknown parametric rule of variability with respect to inhibitory neuron type in a V1 model, where analytic techniques became untenable.
Furthermore, we identified multiple regimes of SC connectivity which confer rapid task switching, and used the structural analytic tools of EPI to gain a mechanistic understanding of circuit responses.
These valuable theoretical insights illustrate the value of deep inference for the interrogation of neural circuit models.

%-------- old sentences -----------

%The inverse problem is crucial for reasoning about likely parameter values, uniquenesses and degeneracies, attractor states and phase transitions, and predictions made by the model.

%This creates an unfavorable tradeoff.  On the one hand, one may tractably analyze systems of equations with unrealistic assumptions (for example symmetry or gaussianity), mathematically formalizing how parameters affect computation in a too-simple model.  
%On the other hand, one may choose a more biologically accurate, scientifically relevant model at the cost of \emph{ad hoc} approaches to analysis (such as simply examining simulated activity), potentially resulting in bad inference of parameters and thus erroneous scientific predictions or conclusions.

%Exact Bayesian inference is rarely tractable, so Markov chain Monte Carlo (MCMC) and variation inference (VI) techniques are used in practice.

% now transition to deep ML...
%Of course, this same tradeoff has been confronted in many scientific fields characterized by the need to do inference in complex models.
%In response, the machine learning community has made remarkable progress in recent years, via the use of deep neural networks as a powerful inference engine: a flexible function family that can map observed phenomena (in this case the measurable signal of some computation) back to probability distributions quantifying the likely parameter configurations. 
%One celebrated example of this approach from machine learning, of which we draw key inspiration for this work, is the variational autoencoder \cite{kingma2013auto, rezende2014stochastic}, which uses a deep neural network to induce an (approximate) posterior distribution on hidden variables in a latent variable model, given data.

%However, these inference tools have not significantly influenced the study of theoretical neuroscience models, for at least three reasons.
%First, at a practical level, the nonlinearities and dynamics of many theoretical models are such that conventional inference tools typically produce a narrow set of insights into these models.
%Indeed, only in the last few years has deep learning research advanced to a point of relevance to this class of problem.
%Second, the object of interest from a theoretical model is not typically data itself, but rather a qualitative phenomenon -- inspection of model behavior, or better, a measurable signature of some computation -- an \emph{emergent property} of the model.  
%Third, because theoreticians work carefully to construct a model that has biological relevance, such a model as a result often does not fit cleanly into the framing of a statistical model.
%Technically, because many such models stipulate a noisy system of differential equations that can only be sampled or realized through forward simulation, they lack the explicit likelihood and priors central to the probabilistic modeling toolkit.

%First, we stipulate a bijective deep neural network that induces a flexible family of probability distributions over model parameterizations with a probability density we can calculate \cite{rezende2015variational, dinh2016density, papamakarios2017masked}.
%Second, we quantify the notion of emergent properties as a set of moment constraints on datasets generated by the model.  
%Thus, an emergent property is not a single data realization, but a phenomenon or a feature of the model, which is ultimately the object of interest in theoretical neuroscience.
%Conditioning on an emergent property requires a variant of deep probabilistic inference methods, which we have previously introduced \cite{loaiza2017maximum}.
%Third,  because we cannot assume the theoretical model has explicit likelihood on data or the emergent property of interest, we use stochastic gradient techniques in the spirit of likelihood free variational inference \cite{tran2017hierarchical}.    
%Taken together, emergent property inference (EPI) provides a methodology for inferring parameter configurations consistent with a particular emergent phenomena in theoretical models.
%We use a classic example of parametric degeneracy in a biological system, the stomatogastric ganglion \cite{goldman2001global}, to motivate and clarify the technical details of EPI.

%First, we use EPI to produce a set of verifiable hypotheses of input-responsivity in a four neuron-type dynamical model of primary visual cortex; we then validate these hypotheses in the model.
%Second, we demonstrated how the systematic application of EPI to levels of task performance can generate experimentally testable hypotheses regarding connectivity in superior colliculus.  
%Third, we use EPI to uncover the sources of error in a low-rank recurrent neural network executing a simple mathematical task.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}
\subsection{Motivating emergent property inference of theoretical models} \label{results_motivating}

%Consideration of the typical workflow of theoretical modeling clarifies the need for emergent property inference.  
Consideration of the typical workflow of theoretical modeling clarifies the need for emergent property inference.  
First, one designs or chooses an existing model that, it is hypothesized, captures the computation of interest. 
To ground this process in a well-known example, consider the stomatogastric ganglion (STG) of crustaceans, a small neural circuit which generates multiple rhythmic muscle activation patterns for digestion \cite{marder2002cellular}.
Despite full knowledge of STG connectivity and a precise characterization of its rhythmic pattern generation, biophysical models of the STG have complicated relationships between circuit parameters and neural activity \cite{goldman2001global, prinz2004similar}.
%A model of the STG \cite{gutierrez2013multiple} is shown schematically in Figure \ref{fig:STG}A, and note that the behavior of this model will be critically dependent on its parameterization -- the choices of conductance parameters $z = [g_{\text{el}}, g_{\text{synA}}]$.
A subcircuit model of the STG \cite{gutierrez2013multiple} is shown schematically in Figure \ref{fig:STG}A, and note that the behavior of this model will be critically dependent on its parameterization -- the choices of conductance parameters $\mathbf{z} = [g_{\text{el}}, g_{\text{synA}}]$.
Specifically, the two fast neurons ($f1$ and $f2$) mutually inhibit one another, and oscillate at a faster frequency than the mutually inhibiting slow neurons ($s1$ and $s2$).  
The hub neuron (hub) couples with either the fast or slow population or both.

%Second, once the model is selected, one defines the emergent property, the measurable signal of scientific interest.  
Second, once the model is selected, one defines the emergent phenomena of scientific interest.
%To continue our running STG example, one such emergent property is the phenomenon of \emph{network syncing} -- in certain parameter regimes, the frequency of the hub neuron matches that of the fast and slow populations at an intermediate frequency.
In the STG example, we are concerned with neural spiking frequency, which emerges from the dynamics of the circuit model \ref{fig:STG}B.
%One such emergent property is the phenomenon of \emph{unified intermediacy} -- in certain parameter regimes, the frequency of all neurons match at an intermediate firing rate.
An interesting emergent property of this stochastic model is when the hub neuron fires at an intermediate frequency between the intrinsic spiking rates of the fast and slow populations.
%This emergent property is shown in Figure \ref{fig:STG}D at a frequency of 0.53Hz.
This emergent property is shown in Figure \ref{fig:STG}C at an average frequency of 0.55Hz.

%Third, qualitative parameter analysis ensues: since precise mathematical analysis is intractable in this model, a brute force sweep of parameters is done \cite{gutierrez2013multiple}.
%Subsequently, a qualitative description is formulated to describe the different parameter configurations that lead to the emergent property.  
%In this last step lies the opportunity for a precise quantification of the emergent property as a statistical feature of the model. 
% Once we have such a methodology, we can infer a probability distribution over parameter configurations that produce this emergent property.
Third, parameter analyses ensue: brute-force parameter sweeps, ABC sampling \cite{liepe2014framework}, and sensitivity analyses \cite{raue2009structural} are all routinely used to reason about what parameter configurations lead to an emergent property.
In this last step lies the opportunity for a precise quantification of the emergent property as a statistical feature of the model.
Once we have such a methodology, we can infer a probability distribution over parameter configurations that produce this emergent property.

%A. For a choice of model (STG) and emergent property (network syncing), emergent property inference (EPI, gray box) learns a distribution of the model parameters $z = \left[g_{\text{el}}, g_{\text{synA}} \right]$ producing network syncing.  
%In the STG model, jagged connections indicate electrical coupling having electrical conductance $g_{\text{el}}$. 
%Other connections in the diagram are inhibitory synaptic projections having strength $g_{\text{synA}}$ onto the hub neuron, and $g_{\text{synB}}=5$nS for mutual inhibitory connections.  
%Network syncing traces are colored by log probability density of their generating parameters (stars) in the EPI-inferred distribution.  
%B. The EPI distribution of STG model parameters producing network syncing.  
%Samples are colored by log probability density.  
%Distribution contours of emergent property value error are shown at levels of $2.5 \times 10^{-5}$, $5 \times 10^{-5}$, $1 \times 10^{-4}$, $2 \times 10^{-4}$, and $4 \times 10^{-4}$ (dark to light gray).  
%Eigenvectors of the Hessian at the mode of the inferred distribution are indicated as $\mathbf{v}_1$ (solid) and $\mathbf{v}_2$ (dashed) with lengths scaled by the square root of the absolute value of their eigenvalues.  
%Simulated activity is shown for three samples (stars). 
%(Inset) Sensitivity of the system with respect to network syncing along all dimensions of parameter space away from the mode.  
%$v_1$ is sensitive to network syncing ($p<10^{-4}$), while $v_2$ is not ($p=0.67$) (see Section \ref{methods_STG}).  
%C. Deep probability distributions map a latent random variable $\mathbf{w}$ through a deep neural network with weights and biases $\bm{\theta}$ to parameters $\mathbf{z} = f_{\bm{\theta}}(\mathbf{w})$ distributed as $q_{\bm{\theta}}(\mathbf{z})$. 
%D. EPI optimization: To learn the EPI distribution $q_{\bm{\theta}}(\mathbf{z})$ of model parameters that produce an emergent property, the emergent property statistics $T(\mathbf{x})$ are set in expectation over model parameter samples $\mathbf{z} \sim q_{\bm{\theta}}(\mathbf{z})$ and model simulations $\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})$ to emergent property values $\bm{\mu}$.
\begin{figure}
\begin{center}
\includegraphics[scale=0.8]{figures/fig1/fig1.pdf}
\end{center}
\caption{\small Emergent property inference (EPI) in the stomatogastric ganglion.  
\textbf{A}. Conductance-based biophysical model of the STG subcircuit.
In the STG model, jagged connections indicate electrical coupling having electrical conductance $g_{\text{el}}$. 
Other connections in the diagram are inhibitory synaptic projections having strength $g_{\text{synA}}$ onto the hub neuron, and $g_{\text{synB}}=5$nS for mutual inhibitory connections.  
Parameters are represented by the vector $\mathbf{z}$ and membrane potentials by the vector $\mathbf{x}$.
The evolution of this model's activity $\mathbf{x}(t)$ is predicated by differential equations.
\textbf{B}. Spiking frequency $\omega(\mathbf{x}; \mathbf{z})$ is an emergent property statistic. 
In this example, spiking frequency is measured from simulated activity of the STG model at parameter choices of $g_{\text{el}} = 4.5$nS and $g_{\text{synA}} = 3$nS.
\textbf{C}. The emergent property of intermediate hub frequency, in which the hub neuron fires at a rate between the fast and slow frequencies.
This emergent property is defined by a mean and variance on the emergent property statistic.
Simulated activity traces are colored by log probability density of their generating parameters in the EPI-inferred distribution (Panel E).
\textbf{D}. For a choice of model and emergent property, emergent property inference (EPI) learns a deep probability distribution of parameters $\mathbf{z}$.
Deep probability distributions map a simple random variable $\mathbf{z}_0$ through a deep neural network with weights and biases $\bm{\theta}$ to parameters $\mathbf{z} = g_{\bm{\theta}}(\mathbf{z}_0)$.
In EPI optimization, stochastic gradient steps in $\bm{\theta}$ are taken such that entropy is maximized, and the emergent property $\mathcal{X}$ is produced.
The EPI posterior distribution is denoted $q_{\bm{\theta}}(\mathbf{z} \mid \mathcal{X})$. 
\textbf{E}. The EPI posterior producing intermediate hub frequency.
Samples are colored by log probability density.  
Distribution contours of average hub neuron frequency from mean of .55 Hz are shown at levels of $.525$, $.53$, ... $.575$ Hz (dark to light gray away from mean).
Eigenvectors of the Hessian at the mode of the inferred distribution are indicated as $\mathbf{v}_1$ (solid) and $\mathbf{v}_2$ (dashed) with lengths scaled by the square root of the absolute value of their eigenvalues.  
%(Inset) Sensitivity of the system with respect to network syncing along all dimensions of parameter space away from the mode.  
\textbf{F} Simulations from parameters in E.  
(Top) The predictive distribution of the posterior obeys the emergent property.
The black and gray dashed lines show the mean and two standard deviations according the emergent property, respectively.
(Bottom) Simulations at the starred parameter values.
 }
 \label{fig:STG}
\end{figure}

%Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG}A gray box) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution shown in Figure \ref{fig:STG}B.
Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG}D) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution  EPI (Fig. \ref{fig:STG}E).  
This distribution -- represented for clarity as samples from the distribution -- is then the most random parameter distribution producing the emergent property. 
In the STG model, this distribution can be specifically queried to reveal the prototypical parameter configuration for intermediate hub frequency (the mode; Figure \ref{fig:STG}E yellow star), and how it decays based on changes away from the mode.
Indeed, samples equidistant from the mode along these EPI-identified dimensions of sensitivity ($v_1$) and degeneracy ($v_2$) (Fig. \ref{fig:STG}E, arrows) agree with error contours (Fig. \ref{fig:STG}E contours) and have diminished or preserved hub frequency, respectively (Fig. \ref{fig:STG}F activity traces) (see Section \ref{methods_STG}).

\subsection{A deep generative modeling approach to emergent property inference} \label{results_dgm}
Emergent property inference (EPI) formalizes the three-step procedure of the previous section with deep probability distributions.
%First, we consider the model as a coupled set of differential (and potentially stochastic) equations \cite{gutierrez2013multiple}.  
First, we consider the model as a coupled set of differential equations \cite{gutierrez2013multiple}.  
In the running STG example, the model activity $\mathbf{x} = \left[ x_{\text{f1}}, x_{\text{f2}}, x_{\text{hub}}, x_{\text{s1}}, x_{\text{s2}} \right]$ is the membrane potential for each neuron, which evolves according to the biophysical conductance-based equation:
\begin{equation} 
\begin{split}
C_m \frac{d\mathbf{x}(t)}{dt} = -h(\mathbf{x}(t); \mathbf{z}) + d\mathbf{B}
 \end{split}
\end{equation} 
where $C_m$=1nF, and $\mathbf{h}$ is a sum of the leak, calcium, potassium, hyperpolarization, electrical, and synaptic currents, all of which have their own complicated dependence on $\mathbf{x}$ and $\mathbf{z} = [g_{\text{el}}, g_{\text{synA}}]$, and $d\mathbf{B}$ is white gaussian noise (see Section \ref{methods_STG}).

%Second, we define the emergent property, which as above is network syncing: oscillation of the entire population at an intermediate frequency of our choosing (Figure \ref{fig:STG}A bottom).
Second, we define the emergent property, which as above is ``intermediate hub  frequency" (Figure \ref{fig:STG}C).
%Quantifying this phenomenon is straightforward: we define network syncing to be that each neuron's spiking frequency -- denoted $\omega_{\text{f1}}(\mathbf{x})$, $\omega_{\text{f2}}(\mathbf{x})$, etc. -- is close to an intermediate frequency of 0.53Hz.  
%Mathematically, we achieve this via constraints on the mean and variance of $\omega_\alpha(\mathbf{x})$ for each neuron $\alpha \in \{ \text{f1}, \text{f2}, \text{hub}, \text{s1}, \text{s2} \}$:
Quantifying this phenomenon is straightforward: we stipulate that the hub neuron's spiking frequency -- denoted $\omega_{\text{hub}}(\mathbf{x})$ is close to a frequency of 0.55Hz.  
%Mathematically, we achieve this via constraints on the mean and variance of the hub neuron spiking frequency.
Mathematically, we achieve this with two constraints: by setting the mean hub frequency of model generated activity to 0.55Hz.
\begin{equation}\label{eq:EP_STG1}
 \mathbb{E}_{\mathbf{z},\mathbf{x}}\begin{bmatrix} \omega_{\text{hub}}(\mathbf{x}; \mathbf{z}) \end{bmatrix}  ~~=~~  \begin{bmatrix} 0.55 \end{bmatrix}
\end{equation}
and requiring that the variance around this mean is small
\begin{equation}\label{eq:EP_STG2}
\text{Var}_{\mathbf{z},\mathbf{x}}\begin{bmatrix} \omega_{\text{hub}}(\mathbf{x}; \mathbf{z}) \end{bmatrix}  ~~=~~  \begin{bmatrix} 0.025^2 \end{bmatrix}.
\end{equation}
This level of variance was chosen to be low enough to exclude the fast and slow frequencies of the two populations, but large enough to allow structural examination of the compatible parameter space.
In general, an emergent property
\begin{equation}\label{eq:EP_STG}
\mathcal{X}:~~ \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2
\end{equation}
defines a collection of datasets with a statistic $f(\mathbf{x}; \mathbf{z})$ (which may be comprised of multiple statistics) and the means $\bm{\mu}$ and variances $\bm{\sigma}^2$ of those statistics over the datasets.
The choice of $\bm{\sigma}^2$ predicates the degree of variability in around mean $\bm{\mu}$ that is consistent with the emergent property.

Third, we perform emergent property inference: we find a distribution over parameter configurations $\mathbf{z}$, and insist that samples from this distribution produce the emergent property; in other words, they obey the constraints introduced in Equation \ref{eq:EP_STG}.  
This distribution will be chosen from a family of probability distributions $\mathcal{Q} = \left\{ q_{\bm{\theta}}(\mathbf{z}) : \bm{\theta} \in \Theta \right\}$, defined by a deep generative distribution of the normalizing flow class \cite{rezende2015variational, dinh2017density, papamakarios2017masked} -- neural networks which transform a simple distribution into a suitably complicated distribution (as is needed here).  
This deep distribution is represented in Figure \ref{fig:STG}D (see Section \ref{methods_EPI}).
Many distributions in $\mathcal{Q}$ will respect the emergent property constraints, so we choose a normative selection principle imposing no additional structure beyond the production of the emergent property \cite{jaynes1957information, elsayed2017structure, loaiza2017maximum, savin2017maximum}, which is the same normative principle of Bayesian inference (see Section \ref{methods_VI}).

>>>>

\linenumbers

\subsection{Scaling inference in RNNs with EPI} \label{results_LRRNN}
Transient amplification is a hallmark of neural activity throughout cortex, and is often thought to be intrinisically generated by recurrent connectivity in the responding cortical area \cite{murphy2009balanced, hennequin2014optimal, bondanelli2019population}.
It has been shown that to generate such amplified, yet stabilized responses, the connectivity of RNNs must be non-normal \cite{goldman2009memory, murphy2009balanced}, and satisfy additional constraints \cite{bondanelli2020coding}.
In theoretical neuroscience, RNNs are optimized and then examined to show how dynamical systems could execute a given computation \cite{sussillo2014neural, barak2017recurrent}, but such biologically realistic constraints on connectivity are  ignored during optimization for practical reasons.
In general, access to distributions of connectivity adhering to theoretical criteria like stable amplification, chaotic fluctuations \cite{sompolinsky1988chaos}, or low tangling \cite{russo2018motor} would add great scientific value and contextualization to existing research with RNNs.
Here, we use EPI to learn RNN connectivities producing stable amplification, and demonstrate the superior scalability and efficiency of EPI to alternative approaches.
\begin{figure}
\begin{center}
\includegraphics[scale=.8]{figures/fig4/fig4.pdf}
\end{center}
\caption{\small 
\textbf{A}. Wall time of EPI (blue), SNPE (orange), and SMC-ABC (green) to converge on RNN connectivites producing stable amplification.  
Each dot shows convergence time for an individual random seed.
For reference, the mean wall time for EPI to achieve its full constraint convergence (means and variances) is shown (blue line).
\textbf{B}. Simulation count of each algorithm to achieve convergence. Same conventions as A.
\textbf{C}. The predictive distributions of connectivities inferred by EPI (blue), SNPE (orange), and SMC-ABC (green), with reference to $\mathbf{x}_0 = \bm{\mu}$ (gray star).
\textbf{D}. Simulations of networks inferred by each method ($\tau=100ms$).  Each trace (15 per algorithm) corresponds to simulation of one $z$.  (Below) Ratio of obtained samples producing stable amplification, monotnoic decay, and instability.
}
\label{fig:LRRNN}
\end{figure}
